Sample variance in weak lensing: how many simulations are required?

### 2016-01-11 Paper/paper.tex
modified Figure 5

%\documentclass[useAMS, usenatbib,usegraphicx,letter]{mn2e}
%\documentclass[11pt]{article}
\documentclass[reprint,aps,prd,superscriptaddress,showkeys,showpacs]{revtex4-1}
\usepackage{epsfig,amsmath,natbib}

@@

Constraining cosmology using weak gravitational lensing consists of
comparing a measured feature vector of dimension $N_b$ with its
simulated counterpart. An accurate estimate of the $N_b\times N_b$
feature covariance matrix $\bb{C}$ is essential to obtain accurate
parameter confidence intervals. When $\bb{C}$ is measured from a set
of simulations, an important question is how large this set should
be. To answer this question, we construct different ensembles of $N_r$
realizations of the shear field, using a common randomization
procedure that recycles the outputs from a smaller number $N_s\leq
N_r$ of independent ray-tracing $N$--body simulations.  We study
parameter confidence intervals as a function of ($N_s,N_r$) in the
range $1\leq N_s\leq 200$ and $1\leq N_r\lesssim 10^5$.  Previous work
\citep{DodelsonSchneider13} has shown that Gaussian noise in the
feature covariance lead, at quadratic order, to an $O(1/N_r)$
degradation of the parameter confidence intervals. Using a variety of
lensing features measured in our simulations, including shear-shear
power spectra and peak counts, we show that cubic and quartic
covariance fluctuations lead to additional $O(1/N_r^2)$ error
degradation that is not negligible when $N_r$ is only a factor of few
larger than $N_b$. We study the large $N_r$ limit, and find that a
single $N$--body simulation ($N_s=1$) can be repeatedly recycled to
produce as many as $N_r=10^5$ mutually independent shear maps. As a
result, a small number of simulations ($N_s=1$ or $2$) is sufficient
to forecast parameter confidence intervals at percent
accuracy.

@@

\begin{figure}
\includegraphics[scale=0.3]{Figures/scaling_nr.eps}
\caption{Bias in the variance of $w$,
  $\langle\h{\sigma}^2_w\rangle-\sigma^2_{w,\infty}$, as a function of
  the number of realizations $N_r$ used to estimate the covariance
  (eq.~\ref{covest}). Results for different ensembles, generated from
  $N_s=$1 (black), 2 (blue), 5 (green), 50 (red), and 100 (purple)
  simulations are shown. The figure shows both the measured trends
  (solid lines) and their best fits based on eq.~(\ref{ourscaling}),
  with $D$ treated as a free parameter.  The thick black curve
  corresponds to an ensemble of $N_r=128000$ realizations from a
  single simulation ($N_s=1$).  No systematic deviation from the
  expected scaling is seen, even in the extreme last case.  }
\label{wvar_nr}
\end{figure}

@@

In Figure \ref{wvar_nr}, we examine how the degradation in the $w$
constraint depends on the number of simulations used to estimate the
covariance, in the limit of large $N_r$. We find excellent agreement
with the expected scaling (eq.~\ref{ourscaling}) up to values
comparable to $N_r\sim 10^5$. We therefore conclude that a single
$N$--body simulation is sufficient to construct an ensemble of
$O(10^5)$ mutually independent convergence power spectra.

@@

\item When the feature covariance matrix is measured from simulations,
  parameter constraints are degraded. This degradation is appreciably
  larger than the $O(1/N_r)$ computed by \citep{DodelsonSchneider13}
  when the number of realizations $N_r$ is only a factor of few larger
  than the feature vector size $N_b$.
\item We can recycle a single $N$--body simulation to produce an
  ensemble of $O(10^5)$ independent maps. The mean feature measured
  from a shear ensemble, though, could be inaccurate if only one
  $N$--body simulation is used.
\item As few as one or two independent $N$--body simulations are
  sufficient to forecast $w$ errorbars to 1\% accuracy, provided that
  a sufficiently large number $N_r$ of realizations are used to
  measure feature covariances.  In particular, provided that biases
  in the inverse covariance are corrected, percent--level forecasts
  require $N_r\gsim 100 (N_b-N_p)$ realizations.
\item Depending on the feature type used to constrain cosmology, a
  larger number of $N$--body simulations might be needed to measure
  accurate ensemble means to an accuracy corresponding 10\% of the
  statistical error. If this accuracy requirement is relaxed to 50\%
  of the statistical error, we find that as low as $N_s=2$ simulations
  are sufficient for the feature types we consider in this work.
\end{itemize}
%
Future extensions of this work should involve extending our analysis
to a larger set of cosmological parameters, and to more general
feature spaces, such as the ones that characterize non--Gaussian
statistics (e.g. including higher moments of the $\kappa$ field,
Minkowski Functionals, and higher-order $\kappa$ correlators). While
our results are highly encouraging, and suggest that a single N-body
simulation can be recycled repeatedly, to produce as many as $10^5$
independent shear power spectra or peak count histograms. In order to
scale our results to large future surveys, such as LSST, it will be
necessary to determine if our findinds hold when challenged by larger
and higher-resolution $N$--body simulations \citep{Qcontinuum}.

@@

\begin{equation}
\label{secondmoment}
\langle\delta\h{\Psi}_I\delta\h{\Psi}_J\rangle = \frac{\nu^2\Psi_I\Psi_J + \nu^2\gamma\Psi_{[I}\Psi_{J]}}{4\gamma^2(\gamma-1)(2\gamma+1)}
\end{equation}

\begin{equation}
\label{thirdmoment}
\langle\delta\h{\Psi}_I\delta\h{\Psi}_J\delta\h{\Psi}_K\rangle = \frac{\nu^3\Psi_{[I}\Psi_J\Psi_{K]}}{8\gamma(\gamma-1)(\gamma-2)(\gamma+1)(2\gamma+1)}
\end{equation}

\begin{equation}
\label{fourthmoment}
\langle\delta\h{\Psi}_I\delta\h{\Psi}_J\delta\h{\Psi}_K\delta\h{\Psi}_L\rangle = \frac{\nu^4(2\gamma^2-5\gamma+9)\Psi_{[I}\Psi_{J]}\Psi_{[K}\Psi_{L]}}{16\gamma(\gamma-1)(\gamma-2)(\gamma-3)(2\gamma-1)(\gamma+1)(2\gamma+1)(2\gamma+3)}.
\end{equation}

\end{widetext}
%
Here the square bracket notation is a shorthand for a symmetrization over pair of indices: for example
\begin{equation}
\Psi_{[I}\Psi_{J]} = \Psi_{i_1j_1}\Psi_{i_2j_2} + \Psi_{i_1j_2}\Psi_{i_2j_1}
\end{equation}
%
Eq. (\ref{firstmoment}) expresses the bias in the $\bbh{\Psi}$
estimator that already appears in the literature~\citep{Hartlap07}. If
we want to use the bias-corrected $\bbh{\Psi}$ estimator (required for
the perturbative expansion of eq.~\ref{estimatorcovariance}), we need
to apply an additional factor of $(2\gamma/\nu)^n$ to
eqs. (\ref{firstmoment}--\ref{fourthmoment}), where $n$ is the order
of the moment up to which we are applying the correction. If we limit
ourselves to computing the expectation value of
eq. (\ref{estimatorcovariance}) up to order $O(1/\nu^2)$, we do not
need to worry about this correction for
eqs. (\ref{thirdmoment}--\ref{fourthmoment}), as the dominant term
here is already $O(1/\nu^2)$. The next step is expanding
eq. (\ref{estimatorcovariance}) in powers of $\delta\bbh{\Psi}$ up to
fourth order: this is easily done:

\begin{widetext}
\begin{equation}
\label{fullpertexpansion}
\h{\Sigma}_\bb{p} = \left(\bb{F}^{-1}+\sum_{n=1}^4\delta\bbh{F}^{-1}_{(n)}\right)\bb{d}'^T_0(\bb{\Psi}+\delta\bbh{\Psi})\bb{C}(\bb{\Psi}+\delta\bbh{\Psi})\bb{d}'_0\left(\bb{F}^{-1}+\sum_{n=1}^4\delta\bbh{F}^{-1}_{(n)}\right).
\end{equation}
\end{widetext}
%
Carrying out the calculations is simpler than it looks: because of the
structure of eq. (\ref{fullpertexpansion}), each term in the expansion
is proportional to $\Sigma_\bb{p}f_a(N_b,N_p)/N_r^a$, where
$f_a(N_b,N_p)$ is a polynomial in $N_b$ and $N_p$. Terms proportional
to $N_b$ arise from index contractions of type $\mathrm{tr}(\bb{\Psi
  C})=N_b$, which come from symmetrization terms of type
$\Psi_{[I}\Psi_{J]}$. Symmetrization factors of type
$\Psi_{[I}\Psi_J\Psi_{K]}$ or $\Psi_{[I}\Psi_{J]}\Psi_{[K}\Psi_{L]}$,
on the other hand, give rise to contractions of type
$\mathrm{tr}(\bb{\Psi
  C})\mathrm{tr}(\bb{\bb{F}\bb{F}^{-1}})=N_bN_p$. Moreover, we know
that, at every order $O(1/N_r^a)$, $f_a(N_b,N_p)$ has to be
proportional to $N_b-N_p$, because it must vanish when $N_b=N_p$. The
reason for this is that if the feature derivative matrix $\bb{d}'_0$
is square and invertible (which it should be in absence of
degeneracies), then eq. (\ref{estimatorcovariance}) reduces to

>>>>

\documentclass[reprint,aps,prd,superscriptaddress,showkeys,showpacs]{revtex4-1}
\usepackage{epsfig,amsmath,natbib}

@@

Constraining cosmology using weak gravitational lensing consists of
comparing a measured feature vector of dimension $N_b$ with its
simulated counterpart. An accurate estimate of the $N_b\times N_b$
feature covariance matrix $\bb{C}$ is essential to obtain accurate
parameter confidence intervals. When $\bb{C}$ is measured from a set
of simulations, an important question is how large this set should
be. To answer this question, we construct different ensembles of $N_r$
realizations of the shear field, using a common randomization
procedure that recycles the outputs from a smaller number $N_s\leq
N_r$ of independent ray-tracing $N$--body simulations.  We study
parameter confidence intervals as a function of ($N_s,N_r$) in the
range $1\leq N_s\leq 200$ and $1\leq N_r\lesssim 10^5$.  Previous work
\citep{DodelsonSchneider13} has shown that Gaussian noise in the
feature covariance lead, at quadratic order, to an $O(1/N_r)$
degradation of the parameter confidence intervals. Using a variety of
lensing features measured in our simulations, including shear-shear
power spectra and peak counts, we show that cubic and quartic
covariance fluctuations lead to additional $O(1/N_r^2)$ error
degradation that is not negligible when $N_r$ is only a factor of few
larger than $N_b$. We study the large $N_r$ limit, and find that a
single $N$--body simulation ($N_s=1$) can be repeatedly recycled to
produce as many as $N_r={\rm few}\times10^4$ mutually independent shear maps. As a
result, a small number of simulations ($N_s=1$ or $2$) is sufficient
to forecast parameter confidence intervals at percent
accuracy.

@@

\begin{figure}
\includegraphics[scale=0.3]{Figures/scaling_nr.eps}
\caption{Bias in the variance of $w$,
  $\langle\h{\sigma}^2_w\rangle-\sigma^2_{w,\infty}$, as a function of
  the number of realizations $N_r$ used to estimate the covariance
  (eq.~\ref{covest}). The figure shows both the measured trends
  (solid lines) and their theoretical behavior according to eq.~(\ref{ourscaling}) with $D=N_b-N_p$. The asymptotic variance $\sigma^2_{w,\infty}$ has been estimated with a linear regression of $\langle\h{\sigma}^2_w\rangle$ vs $1/N_r$ with $500\leq N_r\leq 10000$ (black), with $\langle\h{\sigma}^2_w\rangle(N_r=50000)$ (red) and $\langle\h{\sigma}^2_w\rangle(N_r=100000)$ (green). No deviations from the expected $1/N_r$ behavior are observed until $N_r={\rm few}\times 10^4$.}
\label{wvar_nr}
\end{figure}

@@

In Figure \ref{wvar_nr}, we examine how the degradation in the $w$
constraint depends on the number of simulations used to estimate the
covariance, in the limit of large $N_r$. We find excellent agreement
with the expected scaling (eq.~\ref{ourscaling}) up to values
comparable to $N_r\sim {\rm few}\times10^4$. We can see from the plot that around these values of $N_r$ the $\langle\hat{\sigma}^2_w\rangle - \sigma^2_{w,\infty}$ curve becomes noisy and reaches negative values. This is a clear indication that the $1/N_r$ behavior is broken and a plateau is reached. The negative values in the plot are a consequence of the noise in the estimation of the $\langle\hat{\sigma}^2_w\rangle$ plateau value. We therefore conclude that a single $N$--body simulation is sufficient to construct an ensemble of $O(10^4)$ mutually independent convergence power spectra, before realizations start repeating themselves. For $N_r\gg10^4$, the shear realizations cannot be considered all independent.

@@

\item When the feature covariance matrix is measured from simulations,
  parameter constraints are degraded. This degradation is appreciably
  larger than the $O(1/N_r)$ computed by \citep{DodelsonSchneider13}
  when the number of realizations $N_r$ is only a factor of few larger
  than the feature vector size $N_b$.
\item We can recycle a single $N$--body simulation to produce an
  ensemble of $O(10^4)$ independent maps. The mean feature measured
  from a shear ensemble, though, could be inaccurate if only one
  $N$--body simulation is used.
\item As few as one or two independent $N$--body simulations are
  sufficient to forecast $w$ errorbars to 1\% accuracy, provided that
  a sufficiently large number $N_r$ of realizations are used to
  measure feature covariances.  In particular, provided that biases
  in the inverse covariance are corrected, percent--level forecasts
  require $N_r\gsim 100 (N_b-N_p)$ realizations.
\item Depending on the feature type used to constrain cosmology, a
  larger number of $N$--body simulations might be needed to measure
  accurate ensemble means to an accuracy corresponding 10\% of the
  statistical error. If this accuracy requirement is relaxed to 50\%
  of the statistical error, we find that as low as $N_s=2$ simulations
  are sufficient for the feature types we consider in this work.
\end{itemize}
%
Future extensions of this work should involve extending our analysis
to a larger set of cosmological parameters, and to more general
feature spaces, such as the ones that characterize non--Gaussian
statistics (e.g. including higher moments of the $\kappa$ field,
Minkowski Functionals, and higher-order $\kappa$ correlators). While
our results are highly encouraging, and suggest that a single N-body
simulation can be recycled repeatedly, to produce as many as $10^5$
independent shear power spectra or peak count histograms. In order to
scale our results to large future surveys, such as LSST, it will be
necessary to determine if our findinds hold when challenged by larger
and higher-resolution $N$--body simulations \citep{Qcontinuum}.

@@

\begin{equation}
\label{secondmoment}
\langle\delta\h{\Psi}_I\delta\h{\Psi}_J\rangle = \frac{\nu^2\Psi_I\Psi_J + \nu^2\gamma\Psi_{\{I}\Psi_{J\}}}{4\gamma^2(\gamma-1)(2\gamma+1)}
\end{equation}

\begin{equation}
\label{thirdmoment}
\langle\delta\h{\Psi}_I\delta\h{\Psi}_J\delta\h{\Psi}_K\rangle = \frac{\nu^3\Psi_{\{I}\Psi_J\Psi_{K\}}}{8\gamma(\gamma-1)(\gamma-2)(\gamma+1)(2\gamma+1)}
\end{equation}

\begin{equation}
\label{fourthmoment}
\langle\delta\h{\Psi}_I\delta\h{\Psi}_J\delta\h{\Psi}_K\delta\h{\Psi}_L\rangle = \frac{\nu^4(2\gamma^2-5\gamma+9)\Psi_{\{I}\Psi_{J\}}\Psi_{\{K}\Psi_{L\}}}{16\gamma(\gamma-1)(\gamma-2)(\gamma-3)(2\gamma-1)(\gamma+1)(2\gamma+1)(2\gamma+3)}.
\end{equation}

\end{widetext}
%
Here the curly bracket notation is a shorthand for a symmetrization over pair of indices: for example
\begin{equation}
\Psi_{\{I}\Psi_{J\}} = \Psi_{i_1j_1}\Psi_{i_2j_2} + \Psi_{i_1j_2}\Psi_{i_2j_1}
\end{equation}
%
Eq. (\ref{firstmoment}) expresses the bias in the $\bbh{\Psi}$
estimator that already appears in the literature~\citep{Hartlap07}. If
we want to use the bias-corrected $\bbh{\Psi}$ estimator (required for
the perturbative expansion of eq.~\ref{estimatorcovariance}), we need
to apply an additional factor of $(2\gamma/\nu)^n$ to
eqs. (\ref{firstmoment}--\ref{fourthmoment}), where $n$ is the order
of the moment up to which we are applying the correction. If we limit
ourselves to computing the expectation value of
eq. (\ref{estimatorcovariance}) up to order $O(1/\nu^2)$, we do not
need to worry about this correction for
eqs. (\ref{thirdmoment}--\ref{fourthmoment}), as the dominant term
here is already $O(1/\nu^2)$. The next step is expanding
eq. (\ref{estimatorcovariance}) in powers of $\delta\bbh{\Psi}$ up to
fourth order: this is easily done:

\begin{widetext}
\begin{equation}
\label{fullpertexpansion}
\h{\Sigma}_\bb{p} = \left(\bb{F}^{-1}+\sum_{n=1}^4\delta\bbh{F}^{-1}_{(n)}\right)\bb{d}'^T_0(\bb{\Psi}+\delta\bbh{\Psi})\bb{C}(\bb{\Psi}+\delta\bbh{\Psi})\bb{d}'_0\left(\bb{F}^{-1}+\sum_{n=1}^4\delta\bbh{F}^{-1}_{(n)}\right).
\end{equation}
\end{widetext}
%
Carrying out the calculations is simpler than it looks: because of the
structure of eq. (\ref{fullpertexpansion}), each term in the expansion
is proportional to $\Sigma_\bb{p}f_a(N_b,N_p)/N_r^a$, where
$f_a(N_b,N_p)$ is a polynomial in $N_b$ and $N_p$. Terms proportional
to $N_b$ arise from index contractions of type $\mathrm{tr}(\bb{\Psi
  C})=N_b$, which come from symmetrization terms of type
$\Psi_{\{I}\Psi_{J\}}$. Symmetrization factors of type
$\Psi_{\{I}\Psi_J\Psi_{K\}}$ or $\Psi_{\{I}\Psi_{J\}}\Psi_{\{K}\Psi_{L\}}$,
on the other hand, give rise to contractions of type
$\mathrm{tr}(\bb{\Psi
  C})\mathrm{tr}(\bb{\bb{F}\bb{F}^{-1}})=N_bN_p$. Moreover, we know
that, at every order $O(1/N_r^a)$, $f_a(N_b,N_p)$ has to be
proportional to $N_b-N_p$, because it must vanish when $N_b=N_p$. The
reason for this is that if the feature derivative matrix $\bb{d}'_0$
is square and invertible (which it should be in absence of
degeneracies), then eq. (\ref{estimatorcovariance}) reduces to
